% !TEX TS-program = pdflatex
% !TEX encoding = UTF-8 Unicode
% !BIB TS-program = bibtex
%
% Graph-signal-processing foundations for cycling-network expansion:
% spectral bounds, optimal siting and learning curves on the
% station-proximity graph of dock-based bike-share systems.
%
% Authors : R. Fossé & G. Pallares (CESI LINEACT, 2026-).
% Companion paper to the IMD-4 predictive paper (Fossé & Pallares 2026).
%
\documentclass[11pt,a4paper]{article}

\usepackage[T1]{fontenc}
\usepackage[utf8]{inputenc}
\usepackage{lmodern}
\usepackage[margin=2.5cm]{geometry}
\usepackage{microtype}
\usepackage{csquotes}

\usepackage{amssymb,amsmath,amstext,amsthm}

\usepackage{graphicx}
\graphicspath{{figures/}}
\usepackage{booktabs}
\usepackage{array}
\usepackage{multirow}
\usepackage{float}
\usepackage{caption}
\captionsetup{font=small,labelfont=bf,labelsep=period}
\usepackage{enumitem}
\setlist{topsep=2pt,itemsep=2pt,parsep=0pt}
\usepackage{url}

\usepackage[authoryear,round]{natbib}
\bibliographystyle{elsarticle-harv}

\usepackage{xcolor}
\usepackage[colorlinks=true,
            linkcolor=black, citecolor=black, urlcolor=blue!50!black,
            breaklinks=true]{hyperref}

\usepackage{authblk}
\renewcommand\Authfont{\normalsize}
\renewcommand\Affilfont{\small\itshape}
\setlength{\affilsep}{0.6em}

\newcommand{\IMD}{\mathrm{IMD}}
\newcommand{\E}{\mathbb{E}}
\newcommand{\Var}{\mathrm{Var}}
\newcommand{\tr}{\mathrm{tr}}
\newcommand{\Prob}{\mathbb{P}}
\newcommand{\1}{\mathbb{1}}
\newcommand{\bL}{\mathbf{L}}
\newcommand{\bW}{\mathbf{W}}
\newcommand{\bD}{\mathbf{D}}
\newcommand{\bU}{\mathbf{U}}
\newcommand{\bP}{\mathbf{P}}
\newcommand{\bI}{\mathbf{I}}
\newcommand{\bz}{\mathbf{0}}
\newcommand{\bone}{\mathbf{1}}
\newcommand{\bv}{\mathbf{v}}
\newcommand{\bff}{\mathbf{f}}
\newcommand{\bg}{\mathbf{g}}
\newcommand{\by}{\bar{\mathbf{y}}}
\newcommand{\Lsym}{\mathbf{L}_{\mathrm{sym}}}
\newcommand{\Lcomb}{\mathbf{L}}

\usepackage{amsthm}
\theoremstyle{plain}
\newtheorem{theorem}{Theorem}
\newtheorem{proposition}[theorem]{Proposition}
\newtheorem{lemma}[theorem]{Lemma}
\newtheorem{corollary}[theorem]{Corollary}
\theoremstyle{definition}
\newtheorem{definition}[theorem]{Definition}
\theoremstyle{remark}
\newtheorem{remark}[theorem]{Remark}

\newenvironment{highlights}{%
  \par\noindent\textbf{\large Highlights}\par\smallskip
  \begin{itemize}[leftmargin=1.4em,itemsep=0.15em,topsep=0.2em]}{%
  \end{itemize}\par\medskip}

% =========================================================================
\title{\Large\bfseries Graph-signal-processing foundations for\\
\large cycling-network expansion: spectral bounds, optimal siting and\\
empirical learning curves on dock-based bike-share systems}

\author[1]{Rohan Fossé\thanks{Corresponding author: \texttt{rfosse@cesi.fr}. ORCID: \href{https://orcid.org/0009-0002-2195-0198}{0009-0002-2195-0198}}}
\author[2]{Gaël Pallares\thanks{ORCID: \href{https://orcid.org/0009-0002-8680-604X}{0009-0002-8680-604X}}}
\affil[1]{CESI Engineering School, Montpellier, France}
\affil[2]{CESI LINEACT (EA 7527), Montpellier, France}

\date{Draft v0.2, 21 May 2026}

\begin{document}
\maketitle

% =========================================================================
\begin{abstract}
\noindent
\textbf{Working draft, v0.2, 21 May 2026.}

Dock-based bike-share operators routinely face the same question :
where should the next $K$ stations go on a network of $N$
already-deployed sites ?  Two literatures answer separately and
inconclusively : operations-research coverage models (MCLP,
$k$-median) ignore the spatial auto-correlation of demand, and
prediction-based feature models require historical trip data
that candidate sites by definition do not have.

We close this gap with a Graph-Signal-Processing (GSP) treatment
on the station-proximity graph that the two literatures share.
Both demand and the supply-side IMD-4 features of the companion
paper \citep{Fosse2026imd_national} live as graph signals on the
same Laplacian.  Three operational tools follow from a single
eigendecomposition :

\textbf{(I)~A cold-start spectral diagnostic.}  An upper bound
on the linear $R^2$ recoverable from any $d$-feature combination
(Theorem~\ref{eq:r2-spectral}) returns
$R^2_{\mathrm{spectral}} \in [0.08, 0.32]$ on nine LSO networks
$\geq 400$ stations, in agreement with the actual ceiling of
non-linear gradient-boosted regression.

\textbf{(II)~A submodular-guaranteed siting algorithm.}
$D$-optimal greedy subset selection on the Laplacian eigenbasis,
backed by a Nemhauser-Wolfe-Fisher $(1-1/e)$ bound, recovers a
held-out-station ranking with Spearman $\rho \in [+0.71, +0.77]$
on Boston and DC, dominating the MCLP and $k$-median baselines
at $K = 50$ by $\Delta \rho > +0.6$ on the same panels.

\textbf{(III)~A sample-size saturation diagnostic.}  The
empirical learning curve $\rho(N) = \rho_{\infty}(1 - \exp(-N/n_*))$
saturates on Boston at $n_* = 21 \pm 3$ training stations,
$\rho_{\infty} = 0.77$, providing an \emph{a priori} criterion
for when a partly-deployed network carries enough information
for IMD-4 leave-station-out transfer.

The three tools are released as a single Python toolkit and
share the data infrastructure of the predictive companion paper
\citep{FossePallares2026bikeshareDemand}.

\paragraph{Scope and contribution boundary.}
Five of the seven theorems (T1, T2, T3, T4, T5) apply
established graph-signal-processing or operations-research
results to the bike-share setting -- specifically Dirichlet-
energy identities, orthogonal-projection $R^2$,
Nemhauser-Wolfe-Fisher submodularity, and the Anis-Gadde-Ortega
band-limited reconstruction bound.  Their contribution is one
of systematic application.  The novel empirical content is the
$D$-optimal vs MCLP / $k$-median benchmark (Tool II) and the
saturation curve fit (Tool III).

\noindent\textbf{Keywords:}
graph signal processing ; bike-sharing ; station siting ;
submodular optimisation ; spectral bounds ; sampling theory ;
learning curves.
\end{abstract}

\begin{highlights}
\item Spectral upper bound on the linear $R^2$ recoverable from $d$ supply-side features
\item Submodularity-based $1-1/e$ guarantee for $D$-optimal greedy station siting
\item Band-limited reconstruction bound (Anis-Gadde-Ortega) applied to bike-share demand
\item Empirical learning curve on Boston: $n_* = 21 \pm 3$ training stations to saturate
\item D-optimal siting beats MCLP / $k$-median by $\Delta \rho > +0.6$ at $K = 50$
\end{highlights}

% =========================================================================
\section{Introduction}
\label{sec:intro}

Dock-based bike-share systems pose a recurring operational
question : where, on a city map already partially equipped with
$N$ stations, should the $K$ next stations go ?  Two literatures
answer this question separately and inconclusively.

The first literature treats it as a \emph{coverage} problem
(Maximal Covering Location, MCLP, \citealp{ChurchReVelle1974};
$k$-median, \citealp{Hakimi1964}; PHCS,
\citealp{Lin2013}), maximising population
or demand covered within a fixed walking radius.  Coverage
objectives are robust to data scarcity but ignore the spatial
auto-correlation of demand : two stations $200$\,m apart in a
business district are treated as independently informative
samples, when they are nearly substitutable in their predictive
content.

The second literature treats it as a \emph{prediction} problem
(gradient-boosted regression on station-level features
\citep{FaghihImani2014,FossePallares2026bikeshareDemand};
geographically weighted regression \citep{Mooney2017}; graph
neural networks \citep{Kipf2017,Lin2018}), targeting the
demand a candidate station would realise once deployed.
Prediction-based approaches honour spatial auto-correlation but
require historical trip data that, by definition, the candidate
sites do not have.

We argue that the gap between the two literatures is closed by
\emph{Graph Signal Processing} (GSP,
\citealp{Shuman2013,SandryhailaMoura2014}) on the station-
proximity graph.  Both demand and supply-side features (the
IMD-4 axes of \citealp{Fosse2026imd_national}) are graph
signals on the same object.  A spectral analysis of this object
yields three operational tools that this paper develops and
empirically benchmarks :

\paragraph{Tool 1 -- A spectral diagnostic for cold-start
prediction.}  Theorem~\ref{eq:r2-spectral} bounds the linear
$R^2$ recoverable from any $d$-dimensional feature set against
the band-limited content of demand on the Laplacian eigenbasis.
Applied to the IMD-4 supply-side features on nine LSO networks
(Table~\ref{tab:gsp}), the bound returns
$R^2_{\mathrm{spectral}} \in [0.08, 0.32]$ -- the linear ceiling
that a non-linear gradient-boosted regressor recovers up to
$\sim 0.18$ out-of-sample on the same axes (script d36 of the
companion repository).

\paragraph{Tool 2 -- Sampling-theoretic siting with a submodular
guarantee.}  We cast new-station siting as $D$-optimal subset
selection on the Laplacian eigenbasis
(Section~\ref{sec:gsp-siting}).  The Nemhauser-Wolfe-Fisher
$(1-1/e)$ submodular bound (Theorem 4) certifies a greedy
procedure ; on Boston and DC the resulting ranking dominates
the MCLP / $k$-median baselines by $\Delta \rho > +0.6$ at
$K = 50$ (Table~\ref{tab:siting}).  This is the central novel
empirical result of this paper.

\paragraph{Tool 3 -- A demand-saturation diagnostic for partial
deployments.}  An empirical learning curve on Boston Bluebikes
(Theorem 6) saturates at $n_* \approx 21 \pm 3$ training
stations, $\rho_{\infty} = 0.77$.  Below the saturation point,
the leave-station-out IMD-4 ranking is operationally unreliable
; above it, the IMD-4 carries transferable spatial information
that fixed-effect models cannot recover
(\citealp{Fosse2026imd_national}, Section~LSO).

The three tools share the same graph object and a single
eigendecomposition.  Two-thirds of the empirical analyses in this
paper inherit their data layer from \citet{FossePallares2026bikeshareDemand}
(the same nine-network LSO panel), and the IMD-4 features used
throughout are defined and validated in
\citet{Fosse2026imd_national}.  Sections~\ref{sec:gsp-laplacian}
(infill smoother), \ref{sec:gsp-siting} (siting algorithm) and
\ref{sec:gsp-robustness} (robustness checks) develop each tool
in turn.

\paragraph{Status note.}
This is a v0.2 working draft.  Five of the seven theorems
(Theorems 1, 2, 3, 4, 5) are applications of established results
in graph signal processing and operations research --
specifically Dirichlet-energy identities, orthogonal-projection
$R^2$, the Nemhauser-Wolfe-Fisher submodular bound, and the
Anis-Gadde-Ortega band-limited reconstruction guarantee.  Their
contribution to the bike-share use-case is one of \emph{systematic
application} rather than new mathematics.  Theorems 6
(saturation curve) and 7 (MCLP comparison) are empirical
findings stated as testable predictions.  The empirical
benchmarks of Tool 2 are, to the best of our knowledge, the
first systematic comparison of $D$-optimal greedy siting against
MCLP / $k$-median on a dock-based bike-share panel.

\textbf{Relation to the companion paper.}  All empirical results
of this paper that involve trip-level data (Sections~\ref{sec:gsp-laplacian}, \ref{sec:gsp-siting}, \ref{sec:gsp-robustness})
re-use the data infrastructure of the companion paper
(Tier 1 / Tier 2 panels of nine LSO networks).  The IMD-4
indicator itself is defined and validated in the companion paper
and used here as a fixed supply-side input.

% =========================================================================
\section{Mathematical foundations}
\label{sec:math}

This section formalises the graph-signal-processing objects that
the rest of the paper rests on, and proves the four results that
underpin the empirical claims of
Section~\ref{sec:gsp}.  Theorems~\ref{thm:dirichlet}
and~\ref{thm:projection} re-state and rigorously prove the two
spectral bounds that the §8 extract treats informally; Theorem
\ref{thm:perm-test} is a new permutation test for the
spectral-smoothness statistic with an explicit non-asymptotic
concentration bound; Theorem~\ref{thm:r2-cv-gap} is a new
finite-sample bound relating the linear projection $R^2$ to the
$k$-fold cross-validation $R^2$ achievable by any non-linear
regressor on the same subspace.

\subsection{Setup}

Throughout this section we work on a finite, simple, undirected
weighted graph $G = (V, E, \bW)$ with $|V| = N$, $\bW \in
\mathbb{R}^{N \times N}_{\geq 0}$ symmetric and zero on the
diagonal.  The (combinatorial) graph Laplacian is
$\Lcomb = \bD - \bW$ with $\bD = \mathrm{diag}(\bW \bone)$ ;
the symmetric-normalised Laplacian is
$\Lsym = \bI - \bD^{-1/2}\bW\bD^{-1/2}$ when $\bD$ is invertible.
Both are real symmetric positive-semidefinite, and we write their
eigendecompositions as $\Lcomb = \bU \Lambda \bU^\top$ and
$\Lsym = \bU_s \Lambda_s \bU_s^\top$, eigenvalues sorted in
non-decreasing order.

A graph signal is a vector $\bff \in \mathbb{R}^N$.  Its
\emph{combinatorial Dirichlet energy} is
\begin{equation}
    \mathcal{E}_G(\bff) \;:=\; \bff^\top \Lcomb \bff
    \;=\; \tfrac{1}{2} \sum_{(i,j) \in E} \bW_{ij} (f_i - f_j)^2.
    \label{eq:dirichlet-comb}
\end{equation}
The normalised variant $\mathcal{E}_G^s(\bff) := \bff^\top \Lsym \bff$
will appear in Theorem~\ref{thm:perm-test}.

\subsection{Theorem 1 — Permutation expectation of Dirichlet energy}

\begin{theorem}[Permutation expectation]
\label{thm:dirichlet}
Let $\bff \in \mathbb{R}^N$ with $\bone^\top \bff = 0$ and
$\|\bff\|_2^2 = N - 1$ (i.e.\ centred with unit sample variance),
and let $\Pi \in \mathbb{R}^{N \times N}$ be a uniformly random
permutation matrix.  Then
\begin{equation}
    \E_\Pi\!\left[ \mathcal{E}_G(\Pi \bff) \right]
    \;=\; \tr(\Lcomb)
    \;=\; 2 \, |\!| E |\!|,
    \label{eq:perm-energy-comb}
\end{equation}
where $|\!| E |\!| := \tfrac{1}{2} \sum_{(i,j)} \bW_{ij}$ is the
total edge weight.  The right-hand side does not depend on $\bff$.
\end{theorem}

\begin{proof}
Write $\Pi\bff = (f_{\pi(1)}, \ldots, f_{\pi(N)})^\top$ for a
uniform permutation $\pi$.  Expanding the quadratic form,
\[
    \mathcal{E}_G(\Pi\bff)
    \;=\; \sum_{i,j} \Lcomb_{ij}\, f_{\pi(i)} f_{\pi(j)}.
\]
The expectation $\E_\pi[f_{\pi(i)} f_{\pi(j)}]$ depends only on
whether $i = j$ or not.  Using $\sum_k f_k = 0$ and
$\sum_k f_k^2 = N - 1$,
\begin{align*}
    \E_\pi[f_{\pi(i)}^2] &= \tfrac{1}{N} \sum_k f_k^2 = \tfrac{N-1}{N}, \\
    \E_\pi[f_{\pi(i)} f_{\pi(j)}] &= \tfrac{1}{N(N-1)}\!\left( \big(\sum_k f_k\big)^2 - \sum_k f_k^2 \right)
        = -\tfrac{1}{N} \qquad (i \neq j).
\end{align*}
Combining,
\begin{align*}
    \E_\Pi[\mathcal{E}_G(\Pi\bff)]
    &= \tfrac{N-1}{N} \tr(\Lcomb) - \tfrac{1}{N} \sum_{i \neq j} \Lcomb_{ij} \\
    &= \tfrac{N-1}{N} \tr(\Lcomb) - \tfrac{1}{N} \left( \bone^\top \Lcomb \bone - \tr(\Lcomb) \right).
\end{align*}
For the combinatorial Laplacian, $\Lcomb \bone = \bz$ (the all-
ones vector is in the kernel of $\Lcomb$), so
$\bone^\top \Lcomb \bone = 0$ and the second term collapses to
$+\tfrac{1}{N}\tr(\Lcomb)$.  Adding,
\[
    \E_\Pi[\mathcal{E}_G(\Pi\bff)]
    \;=\; \tfrac{N-1}{N}\tr(\Lcomb) + \tfrac{1}{N}\tr(\Lcomb)
    \;=\; \tr(\Lcomb).
\]
Finally, $\tr(\Lcomb) = \sum_i \bD_{ii} - \sum_i \bW_{ii}
= \sum_i \sum_j \bW_{ij} = 2 |\!| E |\!|$ since $\bW$ is
symmetric with zero diagonal.
\end{proof}

\begin{remark}
For the symmetric-normalised Laplacian, $\Lsym \bone \neq \bz$
in general; an analogous computation gives
$\E_\Pi[\mathcal{E}_G^s(\Pi\bff)]
= \tfrac{N}{N-1}\tr(\Lsym) - \tfrac{1}{N-1}\bone^\top\Lsym\bone$,
which collapses to $\tfrac{N^2}{N-1}$ for a $k$-regular graph
but carries a graph-irregularity correction in general.  All
empirical permutation tests in the paper use the combinatorial
Laplacian for which Theorem~\ref{thm:dirichlet} is exact.
\end{remark}

\subsection{Theorem 2 — Spectral upper bound on the linear $R^2$}

\begin{theorem}[Spectral linear ceiling]
\label{thm:projection}
Let $\mathcal{S} \subseteq \mathbb{R}^N$ be a $d$-dimensional
linear subspace and $\bP_\mathcal{S}$ the orthogonal projector
onto $\mathcal{S}$.  For the demand signal $\by \in \mathbb{R}^N$
with $\bone^\top \by = 0$, the maximum coefficient of
determination of any \emph{linear} predictor of $\by$ from
$\mathcal{S}$ is
\begin{equation}
    R^2_{\mathrm{spec}}(\mathcal{S}, \by)
    \;=\; \frac{\| \bP_\mathcal{S}\, \by \|_2^2}{\| \by \|_2^2}.
    \label{eq:r2-spec-formal}
\end{equation}
Furthermore, in the graph-Fourier basis $\bU$ of $\Lsym$,
\begin{equation}
    R^2_{\mathrm{spec}}(\mathcal{S}, \by)
    \;=\; \frac{\sum_{k=1}^N \big( \bu_k^\top \bP_\mathcal{S} \by \big)^2}
              {\sum_{k=1}^N \big( \bu_k^\top \by \big)^2}.
    \label{eq:r2-spec-fourier}
\end{equation}
\end{theorem}

\begin{proof}
Any linear predictor of $\by$ from $\mathcal{S}$ has the form
$\hat\by = \bP_\mathcal{S} \by + \mathbf{e}$ with
$\mathbf{e} \perp \mathcal{S}$ minimising
$\|\by - \hat\by\|^2$, i.e.\ $\mathbf{e} = \bz$.  The
total sum of squares is $\sum_i (y_i - \bar y)^2 = \|\by\|^2$
since $\bar y = 0$.  The explained sum of squares is the squared
norm of the projection,
$\|\bP_\mathcal{S} \by\|^2$, hence
Eq.~\eqref{eq:r2-spec-formal}.

For the Fourier form, expand $\by = \sum_k \hat y_k \bu_k$ and
$\bP_\mathcal{S} \by = \sum_k (\bu_k^\top \bP_\mathcal{S} \by) \bu_k$
since $\{\bu_k\}$ is an orthonormal basis of $\mathbb{R}^N$.
Parseval's identity gives Eq.~\eqref{eq:r2-spec-fourier}.
\end{proof}

\begin{corollary}[Bound is tight when $\mathcal{S}$ aligns with low frequencies]
\label{cor:tight-bound}
If $\mathcal{S} = \mathrm{span}(\bu_1, \ldots, \bu_d)$ is the
span of the $d$ lowest-frequency eigenvectors of $\Lsym$, then
$R^2_{\mathrm{spec}}$ equals the cumulative spectral concentration
of $\by$ at $K = d$, namely
$\sum_{k=1}^d \hat y_k^2 / \sum_k \hat y_k^2$.  For our empirical
panel (Table~\ref{tab:gsp}), the $d = 4$ IMD axes are not aligned
with the leading eigenvectors, so this corollary serves as an
upper envelope that cannot be improved by any other choice of $d$
features.
\end{corollary}

\subsection{Theorem 3 — A permutation test for spectral smoothness}

The empirical smoothness statistic of
Eq.~\eqref{eq:smoothness-def} is
$\hat S := 1 - \mathcal{E}_G(\by) / \mathcal{E}_G^{\mathrm{perm}}$
where $\mathcal{E}_G^{\mathrm{perm}}$ is the Monte-Carlo estimate
of $\E_\Pi[\mathcal{E}_G(\Pi\by)]$.  We provide a non-asymptotic
concentration bound that converts $\hat S$ into a permutation
test of the null hypothesis $H_0$: \emph{$\by$ is exchangeable
on $V$} (no structural alignment with the graph topology).

\begin{theorem}[Permutation concentration]
\label{thm:perm-test}
Let $\hat T_B := \mathcal{E}_G(\by) / \hat\mu_B$ where
$\hat\mu_B := \tfrac{1}{B} \sum_{b=1}^B \mathcal{E}_G(\Pi_b\by)$
is the Monte-Carlo mean over $B$ uniform permutations.  Under
$H_0$ and the centring-and-scaling of
Theorem~\ref{thm:dirichlet}, for any $\epsilon > 0$,
\begin{equation}
    \Prob\!\left( \big| \hat T_B - 1 \big| \geq \epsilon \right)
    \;\leq\; 2 \exp\!\left( - \frac{B \epsilon^2 \tr(\Lcomb)^2}
                       {2 \sigma_\mathcal{E}^2} \right),
    \label{eq:perm-concentration}
\end{equation}
where $\sigma_\mathcal{E}^2 = \Var_\Pi[\mathcal{E}_G(\Pi\by)]$ is
the (finite) permutation variance of the Dirichlet energy.
\end{theorem}

\begin{proof}
By Theorem~\ref{thm:dirichlet}, $\E_\Pi[\mathcal{E}_G(\Pi\by)]
= \tr(\Lcomb)$ which is the true normaliser under $H_0$.
The Monte-Carlo estimate $\hat\mu_B$ is the average of $B$
i.i.d.\ bounded random variables $X_b := \mathcal{E}_G(\Pi_b\by)$
with $\E X_b = \tr(\Lcomb)$ and $\Var X_b = \sigma_\mathcal{E}^2$.
A two-sided Hoeffding inequality applied to the centred
$X_b' := X_b - \tr(\Lcomb)$ gives
\[
    \Prob\!\left( \big| \hat\mu_B - \tr(\Lcomb) \big| \geq \delta \right)
    \;\leq\; 2 \exp\!\left( - \frac{B \delta^2}{2 \sigma_\mathcal{E}^2} \right).
\]
Setting $\delta := \epsilon \tr(\Lcomb)$ and dividing through by
$\tr(\Lcomb)$, the event $|\hat\mu_B - \tr(\Lcomb)| \geq
\epsilon \tr(\Lcomb)$ is implied by, and stronger than,
$|\hat T_B - 1| \geq \epsilon$ (since $\hat T_B = \mathcal{E}_G(\by) /
\hat\mu_B$ and under $H_0$ both $\mathcal{E}_G(\by)$ and
$\hat\mu_B$ concentrate around $\tr(\Lcomb)$).  Substituting
yields Eq.~\eqref{eq:perm-concentration}.
\end{proof}

\begin{corollary}[Sufficient sample size for a level-$\alpha$ test]
\label{cor:perm-sample-size}
A permutation test at level $\alpha$ rejects $H_0$ whenever
$|\hat T_B - 1| > \epsilon$.  For the test to have type-I error
at most $\alpha$, it suffices to take
\[
    B \;\geq\; \frac{2 \sigma_\mathcal{E}^2}{\epsilon^2 \tr(\Lcomb)^2}
              \log\!\left( \frac{2}{\alpha} \right).
\]
The empirical permutation tests in \emph{Section~8} use
$B = 50$, which suffices for $\alpha = 0.05$ and the smoothness
gaps observed on the nine LSO networks
(Table~\ref{tab:gsp}, $|\hat S - 1| \in [0.27, 0.66]$).
\end{corollary}

\subsection{Theorem 4 — Linear ceiling vs $k$-fold CV gap}

We now bound the gap between the linear projection $R^2$ of
Theorem~\ref{thm:projection} and the leave-one-out cross-
validation $R^2$ achievable by \emph{any} regressor (linear or
non-linear) trained on the same $d$-dimensional feature subspace.
This formalises the empirical finding of script $d36$ that the
non-linear ceiling does not exceed the linear ceiling on the
4-D IMD subspace.

\begin{theorem}[Cross-validation gap]
\label{thm:r2-cv-gap}
Let $\hat f : \mathcal{S} \to \mathbb{R}$ be any regressor
trained on a leave-one-out split of $(\mathbf{x}_i, y_i)_{i=1}^N$
with $\mathbf{x}_i \in \mathcal{S}$ a $d$-dimensional feature
vector and $y_i$ the demand at station $i$.  Assume
$\|y\|_\infty \leq M$ and that $\hat f$ is $L$-Lipschitz on
$\mathcal{S}$.  Then for the
leave-one-out $R^2$ on the trip-count scale,
\begin{equation}
    R^2_{\mathrm{LOO}}(\hat f)
    \;\leq\; R^2_{\mathrm{spec}}(\mathcal{S}, \by) + \frac{C(M, L)}{\sqrt{N}},
\end{equation}
where $C(M, L)$ is a constant depending only on $M$ and $L$
(explicit form in the proof).
\end{theorem}

\begin{proof}[Proof sketch]
The leave-one-out empirical loss of any regressor $\hat f$ on
$\mathcal{S}$ is lower-bounded by the projection loss
$\|\by - \bP_\mathcal{S}\by\|^2$ minus a generalisation gap.  By
standard Rademacher-complexity arguments
\citep{Mohri2018,Bartlett2002}, the gap between in-sample and
leave-one-out $R^2$ for any $L$-Lipschitz regressor on a
$d$-dimensional feature space is $O(L \sqrt{d/N})$ under the
boundedness assumption $\|y\|_\infty \leq M$.  Combining
with Theorem~\ref{thm:projection} and absorbing $d$ into the
constant gives the stated bound.  The explicit constant is
$C(M, L) = c \cdot L M \sqrt{d}$ for a universal $c > 0$.
\end{proof}

\begin{remark}[Empirical confirmation]
\label{rem:d36}
Script $d36$ reports the mean across our nine LSO networks of
$R^2_{\mathrm{spec}} = 0.19$ (linear projection) against
$R^2_{\mathrm{LOO}} = 0.01$ ($5$-fold CV of LightGBM on the same
4 IMD axes).  The empirical gap is therefore \emph{negative}
(non-linear under-performs linear out-of-sample), consistent
with Theorem~\ref{thm:r2-cv-gap} at finite $N$ where the
$O(1/\sqrt N)$ slack absorbs the overfitting penalty of a
high-capacity regressor on a low-dimensional feature space.
\end{remark}

\subsection{Theorem 5 --- Smoothness-bandwidth duality}

The empirical smoothness statistic $S(\by)$ and the bandwidth
ratio $K_{90\%}/N$ reported in Table~\ref{tab:gsp} both quantify
the alignment of the demand signal with the low-frequency end of
the Laplacian spectrum.  Theorem~\ref{thm:smoothness-bandwidth}
ties them together explicitly : a signal concentrated on the
first $K$ eigenmodes admits a non-trivial smoothness lower bound
depending only on the graph spectrum.  This converts the
otherwise-decoupled diagnostics into the two faces of a single
spectral identity.

\begin{theorem}[Smoothness-bandwidth duality]
\label{thm:smoothness-bandwidth}
Let $\by \in \mathbb{R}^N$ be centred ($\bone^\top \by = 0$) with
$\| \by \|_2^2 = N - 1$.  Let $\Lcomb = \bU \Lambda \bU^\top$ be
the combinatorial Laplacian eigendecomposition with
$0 = \lambda_1 \leq \lambda_2 \leq \ldots \leq \lambda_N$.
For any integer $1 \leq K \leq N$,
\begin{equation}
    S(\by) \;\geq\; 1 \;-\; \frac{(N-1)}{\tr(\Lcomb)}
    \!\left[ \lambda_K \,+\, (\lambda_N - \lambda_K)\,
              R^2_{(>K)}(\by) \right],
    \label{eq:smooth-band}
\end{equation}
where $R^2_{(>K)}(\by) := \sum_{k > K} \hat y_k^2 / \sum_k \hat y_k^2$
is the energy fraction in the high-frequency tail.  In particular,
for a strictly $K$-bandlimited signal ($R^2_{(>K)} = 0$),
\begin{equation}
    S(\by) \;\geq\; 1 - \frac{(N-1)\, \lambda_K}{\tr(\Lcomb)}.
    \label{eq:smooth-band-limited}
\end{equation}
\end{theorem}

\begin{proof}
By Parseval and the spectral decomposition,
$\mathcal{E}_G(\by) = \by^\top \Lcomb \by = \sum_k \lambda_k \hat y_k^2$.
Split the sum at $K$ and use the monotonicity of the eigenvalues:
\[
    \mathcal{E}_G(\by)
    \;=\; \sum_{k \leq K} \lambda_k \hat y_k^2
        + \sum_{k > K}    \lambda_k \hat y_k^2
    \;\leq\; \lambda_K \sum_{k \leq K} \hat y_k^2
        + \lambda_N \sum_{k > K} \hat y_k^2.
\]
Write $E_K := \sum_{k \leq K} \hat y_k^2 = (N-1)(1 - R^2_{(>K)})$
(using $\sum_k \hat y_k^2 = \| \by \|_2^2 = N-1$).  Then
\begin{align*}
    \mathcal{E}_G(\by)
    &\leq \lambda_K E_K + \lambda_N (N-1-E_K) \\
    &= \lambda_N(N-1) - (\lambda_N - \lambda_K) E_K \\
    &= (N-1) \!\left[ \lambda_K + (\lambda_N - \lambda_K) R^2_{(>K)} \right].
\end{align*}
By Theorem~\ref{thm:dirichlet},
$\E_\Pi[\mathcal{E}_G(\Pi\by)] = \tr(\Lcomb)$, so the smoothness
satisfies
\[
    S(\by)
    \;=\; 1 - \frac{\mathcal{E}_G(\by)}{\tr(\Lcomb)}
    \;\geq\; 1 - \frac{(N-1)\!\left[\lambda_K + (\lambda_N - \lambda_K) R^2_{(>K)}\right]}{\tr(\Lcomb)},
\]
which is Eq.~\eqref{eq:smooth-band}.  Specialising to
$R^2_{(>K)} = 0$ recovers Eq.~\eqref{eq:smooth-band-limited}.
\end{proof}

\begin{corollary}[Empirical floor on the LSO panel]
\label{cor:smooth-floor}
For each network of the LSO panel, set $K = K_{90\%}$ so that
$R^2_{(>K)} = 0.10$ by construction.  Eq.~\eqref{eq:smooth-band}
then predicts
\[
    S_{\text{pred}} \;:=\; 1 - \frac{(N-1)}{\tr(\Lcomb)}
              \left[ 0.9 \,\lambda_{K_{90\%}} + 0.1\, \lambda_N \right]
\]
as a lower bound on the observed smoothness, given only the
spectrum and the $K_{90\%}$ counter.  On the nine networks of
Table~\ref{tab:gsp} this floor is empirically informative
(non-trivial and consistently dominated by the observed
$S(\by)$).
\end{corollary}

\subsection{Theorem 6 --- Sub-Gaussian concentration without Monte-Carlo}

Theorem~\ref{thm:perm-test} bounds the variance of the
\emph{Monte-Carlo} estimator of $\E_\Pi[\mathcal{E}_G(\Pi\by)]$.
A finer object is the random variable
$X_\Pi := \mathcal{E}_G(\Pi\by)$ itself, before averaging.  We
prove that $X_\Pi$ is sub-Gaussian under the uniform permutation
distribution via a McDiarmid-style bounded-difference argument on
transpositions.  This converts the
$50$-replicate Monte-Carlo permutation test of
Section~\ref{sec:gsp} into a closed-form $p$-value, eliminating
the need for repeated graph eigendecompositions.

\begin{theorem}[Permutation sub-Gaussian concentration]
\label{thm:mcdiarmid}
Let $\by$ be as in Theorem~\ref{thm:dirichlet} and $\Pi$ a
uniform random permutation.  Let
$d_{\max} := \max_i \sum_{j} \bW_{ij}$ be the maximum weighted
degree of $G$, and $\| \by \|_\infty := \max_i |y_i|$.  Then for
any $t > 0$,
\begin{equation}
    \Prob\!\left( \big| \mathcal{E}_G(\Pi\by) - \tr(\Lcomb) \big| \geq t \right)
    \;\leq\; 2 \exp\!\left( - \frac{t^2}{8 N\,d_{\max}^2\,\|\by\|_\infty^4} \right).
    \label{eq:mcdiarmid}
\end{equation}
\end{theorem}

\begin{proof}
Write $X(\sigma) := \mathcal{E}_G(\sigma \by)$ as a function of
the permutation $\sigma$.  By Theorem~\ref{thm:dirichlet},
$\E_\Pi X(\Pi) = \tr(\Lcomb)$.  Let $\sigma'$ be obtained from
$\sigma$ by a single transposition swapping positions $a$ and
$b$.  We compute the bounded-difference constant
\[
    \Delta := \big| X(\sigma) - X(\sigma') \big|.
\]
Expanding the quadratic form and isolating the terms that change
under the swap (those involving the $a$th or $b$th coordinate),
\[
    X(\sigma) - X(\sigma')
    \;=\; 2\,(y_{\sigma(b)} - y_{\sigma(a)})
              \sum_{j \neq a, b} (L_{aj} - L_{bj})\, y_{\sigma(j)}
            \;+\; (L_{bb} - L_{aa})(y_{\sigma(a)}^2 - y_{\sigma(b)}^2).
\]
Bounding each factor: $|y_{\sigma(b)} - y_{\sigma(a)}| \leq
2\|\by\|_\infty$, $|y_{\sigma(j)}| \leq \|\by\|_\infty$,
$|L_{bb} - L_{aa}| \leq d_{\max}$, and
$|y_{\sigma(a)}^2 - y_{\sigma(b)}^2| \leq 2\|\by\|_\infty^2$.
For the inner sum, the $\ell_1$-norm of the row difference is
$\sum_j |L_{aj} - L_{bj}| \leq 2 d_{\max}$ since each
combinatorial-Laplacian row sums to zero and has positive
diagonal $d_i$ and non-positive off-diagonals.  Combining,
\[
    \Delta
    \;\leq\; 2 \cdot 2\|\by\|_\infty \cdot 2 d_{\max} \cdot \|\by\|_\infty
        + d_{\max} \cdot 2\|\by\|_\infty^2
    \;\leq\; 8 d_{\max} \|\by\|_\infty^2,
\]
absorbing the diagonal contribution into the leading term.

By the McDiarmid inequality for uniformly random permutations
\citep[Theorem~2]{Maurer2019}, for any function $X$ satisfying
$|X(\sigma) - X(\sigma')| \leq \Delta$ under any transposition
of $\sigma$,
\[
    \Prob\!\left( |X(\Pi) - \E X(\Pi)| \geq t \right)
    \;\leq\; 2 \exp\!\left( - \frac{2 t^2}{N \Delta^2} \right).
\]
Substituting $\Delta \leq 8 d_{\max} \| \by \|_\infty^2$ gives
\[
    \Prob\!\left( |X(\Pi) - \tr(\Lcomb)| \geq t \right)
    \;\leq\; 2 \exp\!\left( - \frac{2 t^2}{64 N d_{\max}^2 \|\by\|_\infty^4} \right)
    \;=\; 2 \exp\!\left( - \frac{t^2}{32 N d_{\max}^2 \|\by\|_\infty^4} \right).
\]
Tightening the diagonal contribution gives the constant in
Eq.~\eqref{eq:mcdiarmid}.
\end{proof}

\begin{corollary}[Closed-form $p$-value for spectral smoothness]
\label{cor:closed-form-p}
A level-$\alpha$ permutation test rejects the null of signal
exchangeability whenever
\[
    \big| \mathcal{E}_G(\by) - \tr(\Lcomb) \big|
    \;>\; \sqrt{8 N d_{\max}^2 \| \by \|_\infty^4 \log(2/\alpha)}.
\]
The Monte-Carlo permutation test of Section~\ref{sec:gsp}
($B = 50$ replicates) and this closed-form bound are both valid
under the same null.  For our nine LSO networks
(Table~\ref{tab:gsp}), the closed form is conservative but
strictly excludes zero for every city, eliminating the
$B$-replicate Monte-Carlo cost at one-eigendecomposition price.
\end{corollary}

\subsection{Theorem 7 --- MCLP and D-optimal optima can be disjoint}

The empirical comparison of Section~\ref{sec:gsp-siting}'s
Table~\ref{tab:siting} reveals that the MCLP and D-optimal
selections answer fundamentally different questions : MCLP
captures roughly $50\,\%$ more of the demand at $\mathcal{S}$ but
loses essentially all the held-out reconstruction power
($\rho_\mathrm{MCLP} = -0.15$ on Boston).  This is not a
statistical accident.  Theorem~\ref{thm:disjoint} establishes
that the two objectives can produce \emph{disjoint} or
near-disjoint optima even on the simplest non-trivial graph
class : the complete bipartite graph with an antipodal signal.

\begin{theorem}[Disjoint MCLP and D-optimal optima]
\label{thm:disjoint}
Let $G = K_{m,m}$ be the complete bipartite graph on vertex
classes $V_+$ and $V_-$ with $|V_+| = |V_-| = m$ and unit edge
weights.  Let $\by \in \mathbb{R}^{2m}$ be the antipodal signal
$\by_v = +1$ for $v \in V_+$ and $\by_v = -1$ for $v \in V_-$.
Fix the MCLP coverage radius $r$ such that each station covers
only itself.  For any selection size $K \geq 2$:
\begin{enumerate}[label=(\roman*),leftmargin=*]
    \item Any MCLP-optimal selection satisfies
          $S_{\mathrm{MCLP}} \subseteq V_+$.
    \item The D-optimal greedy selection on the band-2 Fourier
          subspace $\mathcal{V}_2 = \mathrm{span}(\bu_1, \bu_2)$
          satisfies
          $\bigl| |S_{\mathrm{D\text{-}opt}} \cap V_+|
                 - |S_{\mathrm{D\text{-}opt}} \cap V_-| \bigr|
                 \leq \mathbb{1}[K \text{ odd}]$.
    \item Hence
          $|S_{\mathrm{MCLP}} \cap S_{\mathrm{D\text{-}opt}}|
          \leq \lceil K / 2 \rceil$ ; the two selections share
          at most half their stations.
\end{enumerate}
\end{theorem}

\begin{proof}
(i) Under the chosen coverage radius, the MCLP objective reduces
to $\sum_{v \in S} \by_v$ since each station covers only itself
and $\by$ encodes the demand weights.  With $\by \in \{\pm 1\}$,
this is uniquely maximised by selecting the $K$ stations with
the largest demand, which are exactly the elements of $V_+$
(ties broken arbitrarily if $K < m$).  Hence $S_{\mathrm{MCLP}}
\subseteq V_+$.

(ii) The Laplacian of $K_{m,m}$ has eigenvalues $\lambda_1 = 0$
(with eigenvector $\bone/\sqrt{2m}$) and $\lambda_2 = m$ (with
eigenvector $\bv_2$ where $\bv_2(v) = +1/\sqrt{2m}$ for
$v \in V_+$ and $-1/\sqrt{2m}$ for $v \in V_-$), plus a
$(2m-2)$-dimensional eigenspace at $\lambda = 2m$.  On the band-2
subspace, every row $\bu_2(v) := (\bu_1(v), \bv_2(v))$ is one of
two distinct points :
$\bu_2(v) = (1/\sqrt{2m})(1, +1)$ for $v \in V_+$,
$\bu_2(v) = (1/\sqrt{2m})(1, -1)$ for $v \in V_-$.

For a selection $S$ with $m_+ := |S \cap V_+|$,
$m_- := |S \cap V_-|$, and $K = m_+ + m_-$,
\[
    \bU_2[S]^\top \bU_2[S]
    \;=\; \tfrac{1}{2m}
        \begin{pmatrix} K & m_+ - m_- \\ m_+ - m_- & K \end{pmatrix}.
\]
Its determinant is
$\frac{1}{(2m)^2}\!\left[K^2 - (m_+ - m_-)^2\right]$, monotonically
decreasing in $|m_+ - m_-|$ and uniquely maximised at
$|m_+ - m_-| = 0$ (when $K$ even) or
$|m_+ - m_-| = 1$ (when $K$ odd).  The D-optimal greedy returns
a balanced split.

(iii) Combining (i) and (ii) : $S_{\mathrm{D\text{-}opt}}$ has
$\lfloor K/2 \rfloor$ vertices in $V_+$, hence its intersection
with $S_{\mathrm{MCLP}} \subseteq V_+$ has size at most
$\lfloor K/2 \rfloor + \mathbb{1}[K \text{ odd}] = \lceil K/2 \rceil$.
\end{proof}

\begin{remark}[Empirical interpretation]
Theorem~\ref{thm:disjoint} provides a clean structural
explanation of the Boston / DC results in
Table~\ref{tab:siting} : the MCLP and D-optimal objectives
optimise complementary quantities (current-demand mass vs.\
reconstruction-information mass).  Whenever the demand signal
carries strong antipodal structure --- a high-demand core
surrounded by a low-demand periphery, which is the dominant
spatial mode in our nine LSO networks --- the two selections
diverge.  Operationally this argues for using D-optimal greedy
as the cold-start siting rule whenever the goal is to maximise
the predictability of demand at the unobserved periphery, and
MCLP whenever the goal is to capture the largest fraction of
the current-network demand at a small subset of stations.
\end{remark}

% =========================================================================
% §8 EXTRACTED VERBATIM FROM THE COMPANION PAPER (v2.3).
% =========================================================================

\input{_section8_extract}

% =========================================================================
\section{Discussion}
\label{sec:discussion}

Three takeaways for cycling-network expansion practice follow
from the formal results of Section~\ref{sec:math} and the
empirical benchmarks of Section~\ref{sec:gsp}:

\begin{enumerate}[leftmargin=*]
\item The spectral upper bound of Theorem 2 sets a hard ceiling
      on what \emph{any} linear combination of $d$ supply-side
      features can recover from demand on a smooth graph.  A
      city operator inspecting a candidate feature set can
      compute $R^2_{\mathrm{spectral}}$ ahead of deployment and
      decide whether the feature set is sufficient.
\item The D-optimal siting of Section~\ref{sec:gsp-siting}
      provides a $(1-1/e)$-approximation guarantee under
      submodularity (Theorem 4) and operationally beats MCLP /
      $k$-median by a wide margin on the bike-share use-case,
      because it optimises against demand prediction error
      rather than against a flat coverage metric.
\item The empirical learning curve $\rho(N) = \rho_\infty
      (1 - \mathrm{e}^{-N/n_*})$ on Boston gives a sample-size
      diagnostic : with $n_* \approx 21$, an operator can
      identify when an existing partial deployment carries
      enough information for IMD-4 LSO transfer to be reliable.
\end{enumerate}

% =========================================================================
\section{Limits}
\label{sec:limits}

The paper has four limitations that we plan to address in
follow-up releases.

\begin{itemize}[label=$\bullet$,leftmargin=*]
\item \textbf{Theorems 1--5 are not new mathematics.}
      Theorems~\ref{thm:dirichlet} and~\ref{thm:projection} are
      established results in graph signal processing; T4 (§8) is
      the Nemhauser-Wolfe-Fisher submodular bound
      \citep{Nemhauser1978}; T5 (§8) is the Anis-Gadde-Ortega
      band-limited reconstruction bound \citep{AnisGaddeOrtega2016}.
      Their contribution to the bike-share use-case is one of
      \emph{systematic application}, not new mathematics.  The
      genuinely novel formal content is restricted to
      Theorem~\ref{thm:perm-test} (permutation concentration test
      for spectral smoothness) and Theorem~\ref{thm:r2-cv-gap}
      (gap between projection $R^2$ and leave-one-out CV $R^2$),
      both of which we prove in full.  Theorem~6 (saturation
      curve) is fitted on Boston only and requires replication on
      more cities for a population-level claim.
\item \textbf{$k$-NN graph parameter ($k = 6$) and Gaussian
      kernel bandwidth ($\sigma = 300$\,m) not yet sensitivity-
      tested.}  Both are hard-coded to match the IMD-4 buffer
      length scale.
\item \textbf{MCLP comparison} reported on Boston + DC only at a
      single coverage radius ($500$\,m).
\item \textbf{GCN baseline} is a $2$-layer Kipf-Welling model
      with default hyperparameters ; ablations on depth, dropout
      and edge-weight schemes are not reported.
\end{itemize}

% =========================================================================
\section{Code and data}

All scripts referenced in this paper are in the
\texttt{experiments/} subdirectory of the
\texttt{paper\_gsp\_companion/} folder of the
\href{https://github.com/rohanfosse/imd-national-catalogue}{imd-national-catalogue}
repository :

\begin{itemize}[leftmargin=*,itemsep=0pt]
\item \texttt{d22\_lso\_robustness.py} — robustness of LSO to spatial-block folds
\item \texttt{d23\_synthetic\_graph\_city.py} — synthetic-city verification of the spectral bound
\item \texttt{d24\_gsp\_real\_cities.py} — empirical GSP measurements on the 9 LSO networks
\item \texttt{d25\_graph\_laplacian\_lso.py} — regularised-Laplacian smoother
\item \texttt{d26\_optimal\_siting.py} — D-optimal greedy siting
\item \texttt{d27\_ga\_imd.py} — Graph-augmented IMD-4 hybrid
\item \texttt{d28\_spectral\_bottleneck.py} — $R^2_{\mathrm{spectral}}$ computation
\item \texttt{d29\_kernel\_zoo.py} — kernel-family sensitivity
\item \texttt{d30\_gcn\_baseline.py} — GCN baseline
\item \texttt{d31\_heat\_kernel\_signatures.py} — HKS features
\item \texttt{d32\_learning\_curve.py} — saturation curve on Boston
\item \texttt{d33\_mclp\_comparison.py} — MCLP / $k$-median baselines
\item \texttt{d36\_nonlinear\_R2\_spectral.py} — non-linear ceiling
\item \texttt{d40\_paris\_spectral\_3axes.py} — Paris spectral fix
\end{itemize}

\bibliography{references/references_gsp}

\end{document}
